Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-07 analysis/paper/paper.Rmd
incorporated some final edits and saved 20220407 draft

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume ~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough is within a migration corridor for several Chinook Salmon populations as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough, an approximately 1.9 kilometer reach of the Salmon River, to be 19,499 in the fall and 10,352 in the spring, corresponding with the peak emigrations of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 61,409 juvenile Chinook Salmon, annually. Using a sensitivity analysis, estimated Northern Pikeminnow predation on juvenile Chinook Salmon was equivalent to 377 (95% CI: 161 - 935) returning adults. Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impact on Chinook Salmon recovery in the Upper Salmon River.

@@

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery efforts [@Friesen1999; @Tabor1993; @Beamesderfer1996]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume ~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$
All mark-recapture abundance estimators include a Chapman correction for small sample sizes [@Chapman1951] and assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or misidentified.

@@

rm(fall_results,
   fall_f30, 
   fall_f40, 
   fall_f50, 
   fall_f60, 
   fall_f70, 
   fall_f80, 
   fall_f90,
   spring_results,
   spring_f30, 
   spring_f40, 
   spring_f50, 
   spring_f60, 
   spring_f70, 
   spring_f80, 
   spring_f90)

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration even after accounting for the potential of delayed mixing of marked fish. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledged that their methods were unsuitable for sampling offshore, noting that water velocity, depth, and irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population; however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and marked fish. Emigration of marked individuals would reduce the marking fraction in the population resulting in an upward bias of abundance estimate, but emigration rates would need to be substantial. Further, the multiple census estimators are more robust to this assumption because the marking fraction is estimated daily. We additionally have no reason to believe that emigration occurs during the two-week window as Deadwater Slough provides favorable habitat to Northern Pikeminnow relative to adjacent reaches. Given the size of the sampling area and the short duration of our surveys, the closed population assumption is likely met. This assumption can also affect results if mortality for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed and fish injured prior to release were retained. Nevertheless, the potential for mortality amongst marked and released fish cannot be discounted.

Next, our abundance estimates assumed that capture and recapture events are random samples of the population. Angling methods often have a size selection bias, thereby limiting abundance estimates to a size range susceptible to angling and resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. Although we recaptured two individual fish during multiple days within a survey (indicating that marked fish were still susceptible to angling), we suspect that marked and released fish may not have immediately mixed thoroughly back into the population. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit “good” fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. For example, if marked fish were only 70% as likely to be caught as unmarked fish, the true abundance is closer to 70% of our abundance estimate. This would additionally account for marked and released fish being “hook shy” for a period of time. The adjusted delayed-mixing Schnabel estimator may account for these biases; however the magnitude of bias is unknown.

The spring abundance estimate was smaller than fall abundance estimates due to the spring CPUE being approximately half of the mean fall CPUE. Although our approach assumed equal capture probabilities between the fall and spring surveys, we believe that higher spring flows may result in lower capture probability of Northern Pikeminnow. Therefore, we believe our spring abundance estimate to be conservative.

@@

Our bioenergetics approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999]. Other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration, and so generalizing energy densities among salmonids may not be problematic. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow’s diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the peak emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during fall months when DSR Chinook Salmon are the dominant prey species available in Deadwater Slough as 1) few to no hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, less than 50% Chinook Salmon in the Northern Pikeminnow’s diet may be more likely. Predation on juveniles from any of the ESA-listed salmonid species is detrimental to their recovery. Future work to quantify and identify juvenile salmonids in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model and to understand impacts to local populations.

Chinook Salmon populations above Deadwater Slough are within the Upper Salmon major population group (MPG) which supports eight independent, extant populations including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. Recovery of the MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and the @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. We estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` juvenile Chinook Salmon may be consumed by Northern Pikeminnow assuming that 60% of their diet is fish and 50% of fish prey are Chinook Salmon. For context, this is approximately `r round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 45416, 2)` times the average annual combined DSR and NRR emigration estimates for brood years 2011 - 2018 at a rotary screw trap located in the lower Lemhi River [@McClure2021; @Feeken2020; @Poole2019], the largest population in the Upper Salmon MPG. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of fish prey are Chinook Salmon, we estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30]), big.mark = ",")` juvenile Chinook Salmon may be consumed which is `r round((total_impacts$juv_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 45416) * 100)`% of the total DSR and NRR emigration for brood years 2011 - 2018 from the Lemhi River. Considering that diet scenarios used in our study are likely conservative assumptions, especially during the fall DSR emigration, the estimated number of Chinook Salmon consumed is substantial.

Spawner abundance is perhaps the most important metric considered in determining a population’s viability and productivity. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough is equivalent to `r round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` adults, annually, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 558) * 100)`% of the total mean spawner escapement for the Upper Salmon MPG in 2017 – 2019 [@Kinzer2020]. In the most conservative scenario, we estimated juvenile consumption equivalent to `r round(total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])` adults, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 558) * 100)`% of recent natural-origin adult Chinook Salmon returns. In the least conservative scenarios, adult equivalents approached recent escapements for the entire Upper Salmon MPG. The sensitivity analysis used some simplifying assumptions. For example, it assumed no juvenile mortality between Deadwater Slough and Lower Granite Dam which is an unrealistic assumption. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including both natural-origin juveniles and hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide for recreational fishing opportunities in the Upper Salmon River. Consequently, reducing predation mortality in Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.

Deadwater Slough is a favorable candidate for management or restoration actions to benefit local Chinook Salmon populations. Three potential management actions could reduce predation at Deadwater Slough: 1) removing or reducing the Dump Creek alluvial fan, 2) a local Northern Pikeminnow bounty program to encourage harvest in Deadwater Slough aimed at reducing the predator population size, and 3) adding structure or cover within Deadwater Slough to provide refuge for juvenile salmonids with the goal or reducing predation rates. Reducing or removing the Dump Creek alluvial fan has the benefit of restoring natural fluvial processes in the Salmon River that likely existed in the reach prior to the formation or increase in size of the alluvial fan related to human activities in upper Dump Creek. Restoring natural processes could reduce Northern Pikeminnow densities to levels similar to upstream and downstream reaches where higher survival and transition probabilities for juvenile salmon have been observed compared to Deadwater Slough [@Axel2015; @Ackerman2018; @Porter2019]. Managers ought also to consider the feasibility and net benefit of restoring fluvial processes relative to potential losses for recreational fishing and bird watching opportunities (https://www.audobon.org/important-bird-areas/deadwater-slough).

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids in the Columbia River [@Winther2020]. A local Northern Pikeminnow bounty program could provide monetary incentive for sportfisherman, in addition to boosting the local economy from lost revenues. Bounty programs could be conducted year-round or seasonally to coincide with peak juvenile outmigration(s). Each approach would require continual/annual effort and harvest to suppress the population size and as such, appropriate funding would be needed to support the program. Finally, a more passive approach could be to provide cover or structure within Deadwater Slough that provides refugia for juvenile salmonids during their emigration. Cover should be appropriately sized to provide concealment to juvenile fishes while reducing access by larger fishes like Northern Pikeminnow. Adding cover could be a cost-effective approach; however, its potential effectiveness is unclear.

Although the origins of Deadwater Slough are somewhat ambiguous [@Reichmuth1985; @USACE1986], it is likely that human activities in Dump Creek appear to have either exacerbated or contributed to its formation and in turn created favorable conditions for Northern Pikeminnow. Among the three surveys, our lowest population abundance estimate was greater than 10,000 Northern Pikeminnow that occupy Deadwater Slough, suggesting a remarkably high density given the size of the area. The slow water velocity and lack of cover for fish also create conditions where juvenile salmonids that are rearing in or emigrating through Deadwater Slough, including Chinook Salmon, are susceptible to predation. We estimate that, at a minimum, Northern Pikeminnow in Deadwater Slough consume greater than `r prettyNum(round(min(total_impacts$juv_Lci), -3), big.mark = ",")` juvenile Chinook Salmon (the lower endpoint of the confidence interval for the most conservative scenario); however, it is likely that this value is closer to `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50], -3), big.mark = ",")`, annually. The magnitude of predation by Northern Pikeminnow is equivalent to `r round((min(total_impacts$adult_N) / 558) * 100)`% and `r round((max(total_impacts$adult_N) / 558) * 100)`% of recent adult escapement. Deadwater Slough predates the ESA-listing of Chinook Salmon populations in the Upper Salmon MPG in the 1990s and is therefore unlikely to be the primary cause for the population’s decline. However, predation by Northern Pikeminnow in Deadwater Slough and elsewhere should be considered among limiting factors hindering recovery efforts of ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

@@

The authors extend much appreciation to the many volunteers that assisted with field efforts including collaborators from the Bureau of Reclamation, Idaho Governor’s Office of Species Conservation, Idaho Department of Fish and Game, and the Lemhi Regional Land Trust, among others. Special thanks to Jared Barker, Brian Hamilton, Tulley Mackey, and Chelsea Welke for their help in the field. We further appreciate the administrative support and guidance from staff at Inter-Fluve. This manuscript benefited from reviews by Sean Gibbs, Mark Roes, and Ian Courter. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018) with assistance from the Idaho Governor’s Office of Species Conservation.

@@

<div id="refs"></div>

@@

<div id="tables"></div>

@@

<div id="figures"></div>

>>>>

Predation on emigrating juvenile salmonids by piscivorous fishes is a widely-studied source of mortality within reservoirs and dam tailraces in the Columbia River basin. Native Northern Pikeminnow have been estimated to consume \~8% of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia rivers, annually. Less is known about the interaction between upstream, river-dwelling Northern Pikeminnow and their impacts on salmonid recovery above the reservoir systems. In this study, we examine the abundance of Northern Pikeminnow in a slow-water reach of the Salmon River, Idaho, known as Deadwater Slough, and consider their potential impacts on local Chinook Salmon populations. The Deadwater Slough is within a migration corridor for several Chinook Salmon populations as well as other Endangered Species Act-listings including steelhead and Sockeye Salmon. We estimated the abundance of Northern Pikeminnow in Deadwater Slough, an approximately 1.9 kilometer reach of the Salmon River, to be 19,499 in the fall and 10,352 in the spring, corresponding with the peak emigrations of juvenile Chinook Salmon. Using these abundance values, we estimated Northern Pikeminnow consumption of juvenile Chinook Salmon. Assuming 60% of the Northern Pikeminnow diet is fish, of which 50% is juvenile Chinook Salmon, we estimated the Northern Pikeminnow population can consume 61,409 juvenile Chinook Salmon, annually. Using a sensitivity analysis, estimated Northern Pikeminnow predation on juvenile Chinook Salmon was equivalent to 377 (95% CI: 161 - 935) returning adults. Given the relatively small size of the Salmon River, the large densities and potential consumption rates of Northern Pikeminnow in Deadwater Slough suggest predation likely has consequential impact on Chinook Salmon recovery in the Upper Salmon River.

@@

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery efforts [@Friesen1999; @Tabor1993; @Beamesderfer1996]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume \~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$

All mark-recapture abundance estimators include a Chapman correction for small sample sizes [@Chapman1951] and assumed: (1) the population is closed (no immigration, emigration, births or deaths during the survey period), (2) all fish have equal chance of being caught in subsequent survey days, (3) marking a fish does not affect its chances of recapture, (4) no loss of marks, and (5) marks are not missed or misidentified.

@@

rm(fall_results,
   fall_f30, 
   fall_f40, 
   fall_f50, 
   fall_f60, 
   fall_f70, 
   fall_f80, 
   fall_f90,
   spring_results,
   spring_f30, 
   spring_f40, 
   spring_f50, 
   spring_f60, 
   spring_f70, 
   spring_f80, 
   spring_f90)
```

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration even after accounting for the potential of delayed mixing of marked fish. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledged that their methods were unsuitable for sampling offshore, noting that water velocity, depth, and irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, the maximum depth at Deadwater Slough was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available throughout the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the basin; additional information on contemporary piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population; however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Emigration of marked individuals could reduce the marking fraction in the population leading to an upward bias in abundance estimates, but emigration rates would need to be substantial. The multiple census estimators, which we report, are more robust to this assumption because the marking fraction is estimated daily. We additionally have little reason to believe that emigration occurs during the two-week survey window owing to favorable habitat for Northern Pikeminnow in Deadwater Slough relative to adjacent reaches. Given the size of the sampling area and the short duration of our surveys, the closed population assumption is likely met. This assumption can likewise affect results if mortality occurs for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed and fish injured prior to release were retained. Nevertheless, the potential for mortality amongst marked and released fish cannot be discounted.

Next, our abundance estimates assumed that capture and recapture events are random samples of the population. Angling methods often have a size selection bias, thereby limiting abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. Two individual fish were recaptured during multiple days within a survey, indicating that marked fish were continually susceptible to angling. We additionally suspect that marked and released fish may not have immediately mixed thoroughly back into the population. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. As an example, if marked fish were only 70% as likely to be caught as unmarked fish, the true abundance is closer to 70% of our abundance estimate. This would additionally account for marked and released fish being "hook shy" for a period of time. The adjusted delayed-mixing Schnabel estimator may account for these biases; however the magnitude of bias is unknown.

The spring abundance estimate was smaller than fall abundance estimates due to the spring CPUE being approximately half of the mean fall CPUE (Table \@ref(tab:summary-tbl)). Although our approach assumed equal capture probabilities between the fall and spring surveys, we believe that higher spring flows may result in lower capture probabilities of Northern Pikeminnow. Therefore, we believe our spring abundance estimate to be conservative.

@@

Our bioenergetics approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999]. Other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration, and so generalizing energy densities among salmonids may not be problematic. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow's diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the peak emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during fall months when DSR Chinook Salmon are the dominant prey species available the Salmon River and Deadwater Slough as 1) few to no hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, less than 50% Chinook Salmon in the Northern Pikeminnow's diet may be more likely. Future work to quantify and identify juvenile salmonids in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model and to understand impacts to local populations.

Chinook Salmon populations above Deadwater Slough are within the Upper Salmon major population group (MPG) which supports eight independent, extant populations including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. Recovery of the MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and the @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. We estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` juvenile Chinook Salmon may be consumed by Northern Pikeminnow assuming that 60% of their diet is fish and 50% of fish prey are Chinook Salmon. For context, this is approximately `r round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 45416, 2)` times the average annual combined DSR and NRR emigration estimates for brood years 2011 - 2018 at a rotary screw trap located in the lower Lemhi River [@McClure2021; @Feeken2020; @Poole2019], the largest population in the Upper Salmon MPG. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of fish prey are Chinook Salmon, we estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30]), big.mark = ",")` juvenile Chinook Salmon may be consumed which is `r round((total_impacts$juv_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 45416) * 100)`% of the total DSR and NRR emigration for brood years 2011 - 2018. Considering that diet scenarios used in our study are likely conservative assumptions, especially during the fall DSR emigration, the estimated number of Chinook Salmon consumed is substantial.

Spawner abundance is perhaps the most important metric considered in determining a population's viability and productivity. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough is equivalent to `r round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` adults, annually, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 558) * 100)`% of the total mean spawner escapement for the Upper Salmon MPG in 2017 -- 2019 [@Kinzer2020]. In the most conservative scenario, we estimated juvenile consumption equivalent to `r round(total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])` adults, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 558) * 100)`% of recent natural-origin adult Chinook Salmon returns. In the least conservative scenarios, adult equivalents approached recent escapements for the entire Upper Salmon MPG. The sensitivity analysis used some simplifying assumptions; for example, it assumed no juvenile mortality between Deadwater Slough and Lower Granite Dam which is an unrealistic assumption. Nevertheless, we found converting juveniles consumed to “adult equivalents” a useful exercise to place results in context with a metric commonly used for recovery.

In additional to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including both natural-origin juveniles and hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that juvenile emigration and adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide for recreational fishing opportunities in the Upper Salmon River. Consequently, reducing predation mortality in Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population. Predation on juveniles from any of the ESA-listed salmonid species is likely detrimental to their recovery.

Deadwater Slough is a favorable candidate for management or restoration actions to benefit local Chinook Salmon populations. Three potential management actions could reduce predation at Deadwater Slough: 1) removing or reducing the Dump Creek alluvial fan, 2) a local Northern Pikeminnow bounty program to encourage harvest in Deadwater Slough aimed at reducing the predator population size, and 3) adding structure or cover within Deadwater Slough to provide refuge for juvenile salmonids to reduce predation rates. Reducing or removing the Dump Creek alluvial fan has the benefit of restoring natural fluvial processes in the Salmon River that likely existed in the reach prior to the formation or increase in size of the alluvial fan. Restoring natural processes could reduce Northern Pikeminnow densities to levels similar to upstream and downstream reaches where higher survival and transition probabilities for juvenile salmon have been observed relative to Deadwater Slough [@Axel2015; @Ackerman2018; @Porter2019]. Managers ought also to consider the feasibility and net benefit of restoring fluvial processes relative to potential losses for recreational fishing and bird watching opportunities (<https://www.audobon.org/important-bird-areas/deadwater-slough>).

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids in the Columbia River [@Winther2020]. A local Northern Pikeminnow bounty program could provide monetary incentive for sportfisherman, in addition to boosting the local economy from lost revenues. Bounty programs could be conducted year-round or seasonally to coincide with peak juvenile outmigration(s). Each approach would require continual/annual effort and harvest to suppress the population size and as such, appropriate funding would be needed to support the program. Lastly, a more passive approach could be to provide cover or structure within Deadwater Slough that provides refugia for juvenile salmonids during their emigration. Cover should be appropriately sized to provide concealment to juvenile fishes while reducing access by larger fishes like Northern Pikeminnow. Adding cover could be a cost-effective approach; however, its potential effectiveness is unclear.

Although the origins of Deadwater Slough are somewhat ambiguous [@Reichmuth1985; @USACE1986], it is likely that human activities in Dump Creek appear to have either exacerbated or contributed to its formation, and in turn, created favorable conditions for Northern Pikeminnow. Among the three surveys, our lowest population abundance estimate was greater than 10,000 Northern Pikeminnow that occupy Deadwater Slough, suggesting a remarkably high density given the size of the area. The slow water velocity and lack of cover for fish also create conditions where juvenile salmonids that are rearing in or emigrating through Deadwater Slough, including Chinook Salmon, are susceptible to predation. We estimate that, at a minimum, Northern Pikeminnow in Deadwater Slough consume greater than `r prettyNum(round(min(total_impacts$juv_Lci), -3), big.mark = ",")` juvenile Chinook Salmon (the lower endpoint of the confidence interval for the most conservative scenario); however, it is likely that this value is closer to `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50], -3), big.mark = ",")`, annually. The magnitude of predation by Northern Pikeminnow is equivalent to `r round((min(total_impacts$adult_N) / 558) * 100)`% and `r round((max(total_impacts$adult_N) / 558) * 100)`% of recent adult escapements. Deadwater Slough predates the ESA-listing of Chinook Salmon populations in the Upper Salmon MPG in the 1990s and is therefore unlikely to be the primary cause for the population's decline. However, predation by Northern Pikeminnow in Deadwater Slough and elsewhere should be considered among limiting factors hindering recovery efforts of ESA-listed Chinook Salmon populations in the Upper Salmon MPG.

@@

The authors extend much appreciation to the many volunteers that assisted with field efforts including collaborators from the Bureau of Reclamation, Idaho Governor's Office of Species Conservation, Idaho Department of Fish and Game, and the Lemhi Regional Land Trust, among others. Special thanks to Jared Barker, Brian Hamilton, Tulley Mackey, and Chelsea Welke for their help in the field. We further appreciate the administrative support and guidance from staff at Inter-Fluve. This manuscript benefited from reviews by Sean Gibbs, Mark Roes, and Ian Courter. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018) with assistance from the Idaho Governor's Office of Species Conservation.

@@

::: {#refs}
:::

@@

::: {#tables}
:::

@@

::: {#figures}
:::
